Supplementary materials: Local adaptation and dispersal evolution interact to drive population response to climate change

### 2018-08-30 Manuscript/Main.tex
Updated the manuscript with the new threshold calculations

\section*{Abstract}
Understanding the interplay between ecological and evolutionary factors is critical for predicting species' responses to anthropogenic climate change and for determining optimal conservation practices. Key parallels between range expansions (e.g. of invasive species) and range shifts in response to climate change allow us to expand current theory for range expansions to include initial spatial population structures and predict species' responses to climate change. Here, we use an individual-based model to explore the role of dispersal evolution, local adaptation, and a population's initial spatial structure in driving the dynamics of climate-induced range shifts. We show that populations with more pronounced initial spatial structure had substantially increased extinction probabilities and that evolution of dispersal ability was insufficient to rescue faltering populations. Rather, a population's fate during climate change was determined by the evolved composition of dispersal phenotypes within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived. Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that the initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.

@@

To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S1 and S2). Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change. This yielded a total of $27$ different scenarios, each explored with $200$ simulations. Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions. Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change. For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.

@@

Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate. While some, individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g. Fig. 1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig. 3a\&b). Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes strongly centered on $0$ (Fig. S5-S7). The similarity in evolved changes to dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions versus became extinct.

Instead, the stable distribution of dispersal phenotypes before the beginning of climate change played a large role in determining the ultimate fate of simulated populations. Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig. S8-S10). The populations that survived climate change were those that happened to make up the upper end of these distributions, containing individuals with high dispersal phenotypes (Fig. 3c\&d). In fact, comparing dispersal phenotypes of successful populations across different parameter combinations revealed a threshold in dispersal phenotypes separating individuals from successful and extinct populations (Fig. S8-S10). This threshold varied directly with the speed of climate change, increasing with more rapid climate change. Across all speeds of climate change, this threshold corresponds to roughly similar probabilities ($12\%$, $9\%$, and $7\%$ at slow, moderate, and fast speeds respectively) of an individual keeping pace with climate change (i.e. moving the correct number of patches in the $x$ dimension to track the changing environmental conditions).

@@

\section*{Figure legends}
\paragraph{Figure 1.} A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge. Information on the (a) abundance, (b) dispersal, and (c) fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change. Log transformed mean dispersal phenotypes (b) are shown for each patch. Average patch fitness (c) was calculated based on the mean local environmental niche phenotype and the optimal niche value for that location.

@@

\paragraph{Figure 3.} Patterns in the evolution and the initial distribution of the dispersal trait, highlighting extant simulations. Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Initial distributions of the dispersal trait (c and d) are shown with histograms of the log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions. In all panels, values associated with extant populations are shown in dark blue. Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations). Full results for all parameter combinations are provided in the online supplement.

>>>>

\section*{Abstract}
Understanding the interplay between ecological and evolutionary factors is critical for predicting species' responses to anthropogenic climate change and for determining optimal conservation practices. Key parallels between range expansions (e.g. of invasive species) and range shifts in response to climate change allow us to draw upon current theory for range expansions to include initial spatial population structures and predict species' responses to climate change. Here, we use an individual-based model to explore the role of dispersal evolution, local adaptation, and a population's initial spatial structure in driving the dynamics of climate-induced range shifts. We show that populations with more pronounced initial spatial structure had substantially increased extinction probabilities and that evolution of dispersal ability was insufficient to rescue faltering populations. Rather, a population's fate during climate change was determined by the evolved composition of dispersal phenotypes within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived. Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that the initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.

@@

To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S1 and S2). Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change. This yielded a total of $27$ different scenarios, each explored with $200$ simulations. Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions. Figure 1 shows an example of a single simulation undergoing climate change. For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.

@@

Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate. While some, individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g. Fig. 1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig. 3a\&b). Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes centered on $0$ (Fig. S5-S7). The similarity in evolved changes to dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.

Instead, the stable distribution of dispersal phenotypes before the beginning of climate change played a large role in determining the ultimate fate of simulated populations. Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig. S8-S10). The populations that survived climate change were those that happened to make up the upper end of these distributions, containing individuals with high dispersal phenotypes (Fig. 3c\&d). Comparing the dispersal phenotypes making up successful populations across different parameter combinations revealed a common threshold separating individuals from successful and extinct populations at each speed of climate change (Fig. S8-S10). In fact, this threshold corresponds to the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change. To show this, we used the well known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, where $r$ is the intrinsic growth rate of the population and $D$ is the diffusion coefficient. The $2$ dimensional diffusion coefficient can be calculated directly from the dispersal phenotypes as $D=\frac{1}{4}d^{2}$ since $d^{2}$ represents the mean squared displacement of an individual with phenotype $d$ during a dispersal period (see the model description in the supplemental materials). Using these relationships and solving for the dispersal phenotype producing each of the speeds of climate change used in our simulations yields the threshold value indicated on Figures S8-S10 which distinguishes those populations that ultimately survived climate change from those that went extinct.

@@

\section*{Figure legends}
\paragraph{Figure 1.} A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge. Information on the (a) abundance, (b) dispersal, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change. Log transformed mean dispersal phenotypes (b) are shown for each patch. Average patch fitness (c) was calculated based on the mean local environmental niche phenotype and the optimal niche value for that location.

@@

\paragraph{Figure 3.} Patterns in the evolution and the initial distribution of the dispersal trait, highlighting extant simulations. Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Initial distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions. In all panels, values associated with extant populations are shown in dark blue. Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations). Full results for all parameter combinations are provided in the online supplement.

### 2018-08-30 Manuscript/Supp.tex
Updated the manuscript with the new threshold calculations

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
    \hline
    Habitat gradient at range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
     & None & $0.0025$ & $250$ & $0$ \\
    Shallow & Low & $0.0025$ & $250$ & $0.004$ \\
     & High & $0.0025$ & $250$ & $0.008$ \\ \hline
     & None & $0.025$ & $421.479$ & $0$ \\
    Moderate & Low & $0.025$ & $421.479$ & $0.004$ \\
     & High & $0.025$ & $421.479$ & $0.008$ \\ \hline
     & None & $0.25$ & $421.48$ & $0$ \\
    Stark & Low & $0.25$ & $421.48$ & $0.004$ \\
     & High & $0.25$ & $421.48$ & $0.008$ \\ 
    \hline
  \end{tabular}
\caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios.}
\label{table:Scenarios}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a slow speed of climate change.}
\label{fig:InitDispSlow}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a moderate speed of climate change.}
\label{fig:InitDispMain}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a fast speed of climate change.}
\label{fig:InitDispFast}
\end{figure}

>>>>

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
    \hline
    Habitat gradient at the range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
     & None & $0.0025$ & $250$ & $0$ \\
    Shallow & Low & $0.0025$ & $250$ & $0.004$ \\
     & High & $0.0025$ & $250$ & $0.008$ \\ \hline
     & None & $0.025$ & $421.479$ & $0$ \\
    Moderate & Low & $0.025$ & $421.479$ & $0.004$ \\
     & High & $0.025$ & $421.479$ & $0.008$ \\ \hline
     & None & $0.25$ & $421.48$ & $0$ \\
    Stark & Low & $0.25$ & $421.48$ & $0.004$ \\
     & High & $0.25$ & $421.48$ & $0.008$ \\ 
    \hline
  \end{tabular}
\caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios.}
\label{table:Scenarios}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. The dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change is indicated by the dashed red lines. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a slow speed of climate change.}
\label{fig:InitDispSlow}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. The dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change is indicated by the dashed red lines. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a moderate speed of climate change.}
\label{fig:InitDispMain}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. The dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change is indicated by the dashed red lines. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a fast speed of climate change.}
\label{fig:InitDispFast}
\end{figure}
